Examining Programming Students’ Interaction with Generative AI: A Metacognitive Perspective

### 2024-10-09 CSEDU2025/Paper.tex
perspective interleaving

\abstract{
This study examines the interactions of programming students with a chatbot,
CharlieBot, powered by OpenAI's GPT-3.5 and enhanced with the
Retrieval-Augmented Generation (RAG) technique, within the context of a Java
programming course. It investigates how students use the bot while employing two
metacognitive strategies: interleaving and spacing. Using the logs of student
interactions with the bot, the methodology involved classifying and analyzing
student prompts into eight distinct categories. The findings reveal that while
most interactions emphasize active learning strategies, students demonstrate
limited conscious application of interleaving and spacing. This study
underscores the necessity for increased guidance on the strategic use of AI
tools to enhance learning outcomes.
}

@@

Recent advancements in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming \citep{Peng23}
\citep{Pandey24} and education \citep{Puryear22}. In programming courses,
students can employ GenAI tools to enhance their understanding, receive
personalized feedback, and access detailed explanations. For example, GitHub
Copilot, a tool that is integrated into development environments, assists by
providing real-time suggestions and accelerating code writing, which could help
streamline the learning process \citep{Denny23}. Educational chatbots, on the
other hand, try to guide students through coding challenges, answer questions,
and attempt to foster a deeper comprehension of programming concepts
\citep{Labadze23}.

The study guided by \cite{chan23} revealed that both undergraduate and
postgraduate students exhibit positive attitudes towards the use of GenAI in
teaching and learning, noting that it may improve the depth of their thinking
and understanding. Furthermore, a systematic review conducted by \cite{Lo24}
demonstrated that students could effectively learn from ChatGPT, leading to
improved understanding and academic achievement \citep{Callejo24}. Besides that,
it was also observed that ChatGPT enables students to regulate their learning
pace \citep{Baha24} and can potentially support students’ self-regulated
learning, especially for those with previous technical and disciplinary
knowledge \citep{Xia23}.

However, researchers are concerned about the impact of these tools on
students. The systematic review conducted by \cite{Murillo23} indicated that
the use of ChatGPT could lead to an overreliance on the tool. \cite{chan23} also
noted that this dependency could result in a decrease in critical thinking, as
students could make decisions based solely on the information provided by
ChatGPT.

The student’s confidence is justified, as demonstrated by \cite{Puryear22}, who
found that GitHub Copilot can generate solutions for student assignments with
accuracy rates ranging from 68\% to 95\%. However, this also raises concerns
that students may become overly dependent on GenAI tools, potentially bypassing
the opportunity to develop a deeper understanding of the underlying concepts.
\cite{cai23} further highlighted the overdependence and diminished intellectual
engagement as significant drawbacks associated with the use of ChatGPT in
learning.

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} revealed that at least 54.5\% of students are already using
GenAI for homework. This highlights the need to increase the understanding of
how students interact with these tools and how they can be used to
enhance learning. Furthermore, a review organized by \cite{Lo24} emphasized the
need for extended studies and objective measures to gather more robust
evidence on the use of GenAI tools in education.

In response to the growing need for deeper insights into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational bot powered by gpt-3.5 from OpenAI, enhanced with
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of the metacognitive strategies
employed by the students. To achieve this objective, we formulate three research
questions:

@@

% % What is its main limitation?

@@

>>>>

\abstract{
This study examines the interactions of programming students with a chatbot,
CharlieBot, powered by OpenAI's GPT-3.5 and enhanced with the
Retrieval-Augmented Generation (RAG) technique, within the context of a Java
programming course. It investigates how students use the bot while employing two
metacognitive strategies: interleaving and spacing. Using the logs of student
interactions with the bot, the methodology involved classifying and analyzing
student prompts into eight distinct categories. The findings reveal that while
most interactions emphasize active learning strategies, students demonstrate
limited conscious application of interleaving and spacing. This study
underscores the need for more guidance on the strategic use of AI
tools to improve learning outcomes.
}

@@

Recent advances in Generative Artificial Intelligence (GenAI) have created
new opportunities in both professional programming \citep{Peng23}
\citep{Pandey24} and education \citep{Puryear22}. In programming courses,
students can employ GenAI tools to improve their understanding, receive
personalized feedback, and access detailed explanations. For example, GitHub
Copilot, a tool that is integrated into development environments, assists in
providing real-time suggestions and accelerating code writing, which could help
streamline the learning process \citep{Denny23}. Educational chatbots, on the
other hand, try to guide students through coding challenges, answer questions,
and try to foster a deeper understanding of programming concepts
\citep{Labadze23}.

The study conducted by \cite{chan23} revealed that both undergraduate and
postgraduate students have positive attitudes towards the use of GenAI.
Furthermore, a systematic review organized by \cite{Lo24} demonstrated that
students can effectively learn from ChatGPT, resulting in improved comprehension
and academic achievement \citep{Callejo24}. Furthermore, it was observed that
ChatGPT allows students to regulate their learning pace \citep{Baha24} and can
potentially support self-regulated learning, particularly for students with
prior technical and disciplinary knowledge \citep{Xia23}.

However, researchers have also expressed concerns about the impact of these
tools on students. The systematic review by \cite{Murillo23} indicated that the
use of ChatGPT could lead to overreliance on the tool. \cite{chan23} noted that
overdependence on ChatGPT could lead to a decrease in critical thinking, as
students can make decisions based solely on the information provided by the tool.
As a potential consequence, \cite{Eason23} observed in a study that when
programming students lost access to ChatGPT-4, those who had previously relied
on it saw their performance drop by 17\%. In contrast, students who had never
used the tool were unaffected and outperformed their peers.

The confidence of students in these tools is well founded, as shown by
\cite{Puryear22}, who found that GitHub Copilot can generate solutions for
student assignments with precision rates ranging from 68\% to 95\%. However,
\cite{Boudouaia24} observed that using ChatGPT without a structured learning
approach did not provide significant improvement over traditional self-directed
learning methods in programming. This raises concerns that students may become
overly dependent on generative AI tools, potentially hindering their learning
progress and missing opportunities to develop a deeper understanding of
fundamental concepts.

Regardless of teachers' preferences or beliefs, preliminary surveys conducted
by \cite{Dickey24} indicate that more than 54,5\% of students are already using
GenAI for homework, likely due to their perception of this technology as
advantageous and intuitive \citep{Boudouaia24}. All of these studies highlight
the need to increase the understanding of how students interact with these tools
and how they can be used to improve learning. Furthermore, a review organized by
\cite{Lo24} emphasized the need for extended studies and objective measures to
gather more robust evidence on the use of GenAI tools in education.

In response to the growing need for deeper insight into the use of GenAI tools
in educational settings, this study aims to help address this gap by analyzing
the interactions between students in a Java programming course and an
educational bot powered by gpt-3.5 from OpenAI, enhanced with the
Retrieval-Augmented Generation (RAG) technique. Specifically, it focuses on
examining these interactions within the context of the metacognitive strategies
employed by students. To achieve this objective, we formulate three research
questions:

@@

% What is its main limitation?

@@

In this study, we propose a variation of the interleaving concept, which we call
Perspective Interleaving. Unlike traditional interleaving, which typically
involves alternating between different topics or problem types, Perspective
Interleaving focuses on students' ability to shift between distinct viewpoints.
There are factors that suggest the Perspective Interleaving hypothesis can
benefit student learning. Even without an explicit alternation between different
types of problems, the student can constantly change their perspective, which
relates to the concept of contextual discrimination. Furthermore, Kolb's
Experiential Learning Theory reinforces this idea, stating that alternating
between concrete experiences and abstract reflections is essential to sustain
the learning cycle. Finally, deliberately shifting one's perspective on a
specific problem can help students reflect on their approach and improve their
metacognitive strategies.
